A Neural, Parallel Theorem Prover*

### 2019-05-14 papers/FroCoS19/paper.tex
superficial presentation issues and make it fit

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{mathpartir}
\usepackage{subcaption}
\usepackage{textcomp}
\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neural, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567}\inst{1} \and
	Giles Reger\inst{2}
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}\and
	University of Manchester, UK\\
	\email{regerg@cs.man.ac.uk}
}
\maketitle

@@

\section{Introduction}
Recent advances~ in neural network systems allow for processing graph-structured data in a neural context.
Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs), suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
However, in practice there are several implementation issues~\cite{lerna-discussion} which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
\begin{enumerate}
	\item Proof state in such systems may be very large, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate.
	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
\end{enumerate}
%
Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures. Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} takes an alternative step toward useful neural automatic theorem provers.

\section{Background}
An understanding of neural networks~\cite{neural-survey}, and of first-order logic and automatic theorem provers~\cite{fol-atp-survey} is assumed.
The particular syntax of first-order logic used in this work is presented here as clarification.

\subsection{Syntax of First-Order Logic}
Terms \(t\) are either
\begin{enumerate}
	\item Variables \(x\), bound by a quantifier.
	\item Function applications \(f\left(t_1, t_2, \ldots, t_n\right)\) where \(f\) is an \(n\)-ary function symbol. Constants \(c\) are a degenerate case in which \(n = 0\).
\end{enumerate}
%
Formulae \(\phi, \psi\) are one of
\begin{enumerate}
	\item Truth \(\top\) or falsehood \(\bot\).
	\item Equalities \(t_1 = t_2\).
	\item Predicate applications \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol.
	\item Negations \(\lnot \phi\), implications \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
	\item Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
	\item Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).
\end{enumerate}

\subsection{Modern Automatic Theorem Provers}
State-of-the-art ATP systems (e.g. E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}, CVC4~\cite{CVC4}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.

\subsection{Machine Learning and Theorem Proving}
Despite the efficiency of such systems, they can still spend time exploring areas that a human mathematician would discard, and tuning such systems manually is in general extremely difficult~\cite{portfolio}.
There is therefore interest in applying machine-learning techniques to automatic theorem provers, with the eventual aim of an ``intelligent'' theorem prover which can learn from past experience to develop an intuition, discard uninteresting search space, and tune performance in a more principled way.

Work has proceeded in the fields of premise selection~\cite{DeepMath,MaSh,MaLARea,formula-graph}, static strategy selection~\cite{Bridge,MaLeS,E-MaLeS}, dynamic (run-time) strategy selection~\cite{dynamic-strategy-priority} and more recently, direct proof guidance~\cite{MaLeCoP,FEMaLeCoP,rlCoP,ENIGMA,DNGPS}.
Proof guidance typically involves some form of machine-learned \emph{heuristic} which biases proof search in some way, allowing the prover to avoid parts of the search space deemed uninteresting by the heuristic.

Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.
rlCoP~\cite{rlCoP} is a connection-based reinforcement-learning prover which is not presently neurally-guided, but takes a similar approach to that taken in this paper and achieves impressive results.

\subsection{Neural Networks for Formulae}
Neural networks are well-known tools for supervised learning~\cite{neural-survey}, and combined with trainable convolution/pooling operators are suitable for processing large-scale data such as images~\cite{cnn}.

@@

\subsection{The MPTP Problem Set}
For training and evaluation purposes a set of problems exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
Urban et al.~\cite{rlCoP} took a subset\footnote{\url{https://github.com/JUrban/deepmath/blob/master/M2k_list}} of the large \mizarlarge{} problem set (containing 32,524 problems) and called it \mizarsmall{} (containing 2004 problems).

\section{Design}
\label{section:Design}
In order to achieve the goal of a neural theorem prover without the disadvantages associated with neural approaches, a new design of theorem prover is required.
Popular calculi used in existing ATPs tend to be unfriendly to neural guidance.
For such a system, we require the following from the calculus:
\begin{enumerate}
        \item \emph{Proof state must be small.} Attempting to evaluate large proof states structurally requires a lot of computation and resources. Saturation-based provers can have very large proof states, for example.
        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms tend to operate more efficiently in batches. Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space, without e.g. unifiers. Otherwise, the learning system is trying to learn while blind to the context of the search.
	\item \emph{Subgoals must be intelligible.} Adding ``noise'' such as clausification obscures the original intuition behind a goal, at least for human observers. While this is not necessarily the case for machine-learning algorithms, it seems likely that removing structure and adding artefacts will reduce model performance.
\end{enumerate}
We therefore implement a refutation prover based on a first-order tableaux calculus without unification, on non-clausal formulae.
Each goal in this case is the set of formulae present on the tableau branch.
In this context, proof state is small (only the current branch), evaluation of states is possible in parallel, each branch is independent and contains all information required, and all available structure from the original problem is kept.

\subsection{Search}
In the calculus (see Section \ref{section:Calculus}) for this prover, there are two branching factors: each goal has a set of possible inferences, and each inference contains a set of possible sub-goals.
To prove a goal, at least one inference must be proved.
To prove an inference, all the inferences' sub-goals must be proved.
A simple optimisation observation is that sub-goals might well be shared between inferences from all over the search space, so search becomes a directed acyclic graph, alternating between goals and inferences.

\begin{figure}
	\centering
	\includegraphics[width=0.6\linewidth]{search}
	\caption{Search in the \lerna{} system, showing shared sub-goals.}
\end{figure}

Now the search graph can be explored: in each step, a leaf (goal) node is selected for expansion, and all resulting inferences and sub-goals are added to the graph.
If a goal has no possible inferences, it is satisfiable and can be removed from the search space.
On the other hand, if a goal is trivial (i.e. contains a contradiction), it is unsatisfiable and can be marked as proven.
This idea can be lifted to inferences: if an inference contains any satisfiable sub-goal, it too is satisfiable, whereas if an inference contains all unsatisfiable goals, it is unsatisfiable.
Proof search continues until the timeout is reached or the root goal is shown to be (un-)satisfiable.
In order to dispatch trivial sub-goals quickly, an existing fast \emph{oracle} ATP is used (see Section \ref{section:oracle}).
This may mark goals as (un-)satisfiable, at which point no further exploration is required.

@@

\subsection{Architecture}
The system aims to consume all available CPU and GPU resources as efficiently as possible.
To that end, proof search is asynchronous: the search algorithm generates new sub-goals, which are placed on two separate queues: one for the oracle ATP, another for heuristic evaluation.
Proof search then continues elsewhere, while the oracle ATP is called in parallel on each sub-goal (consuming all available CPU) while the heuristic consumes batches of subgoals, efficiently utilising the available GPU resource.
As information flows backwards from these processes, the search process updates its information about a given sub-goal and propagates that information upwards to the sub-goal's parent inferences, to influence future proof search: see Figure \ref{figure:architecture}.

\begin{figure}
	\includegraphics[width=\linewidth]{architecture}
	\caption{A cartoon showing information flow in the system.}
	\label{figure:architecture}
\end{figure}

\subsection{Implementation}
The Rust~\cite{rust} programming language was chosen to implement the main prover (minus the heuristic), for its memory and thread-safety guarantees, as well as its support for managing memory allocation and algebraic datatypes.
The implementation is currently just under 3,000 lines of Rust, not including \z3{}, the TPTP format parser or the implementation of perfect sharing.
Python 3~\cite{python3} was used for the heuristic due to the large number of libraries available for neural network implementation in Python.
The heuristic is implemented as a server, communicating with the main prover via a TCP socket.
In principle this allows for the heuristic to be a shared resource with a centralised heuristic server, or a load-balanced cluster.

\section{Calculus}
\label{section:Calculus}
The proof calculus used in this context may be extremely general: in fact, any function from goals to a finite set of possible inferences (themselves finite sets of sub-goals) will suffice, as long as each goal remains independent of any other such that the heuristic function can process all available information.  If the inference system is complete, there are no additional constraints such as orderings or fairness to ensure the completeness of the prover, as the balanced search algorithm (see Section \ref{section:Design}) will ensure this.

\lerna{} presently implements a refutation tableaux calculus~\cite{handbook-of-automated-reasoning} without unification.
The calculus described is deliberately na\"ive in order to easily satisfy the design constraints described, but this is a weak point of this work and will be improved in future.
A na\"ive calculus is not necessarily a problem in this context, as the heuristic should in principle select promising areas to explore and ignore uninteresting sub-goals.
However, a more efficient calculus would improve performance where the heuristic fails.

\subsection{Refutation Tableaux}
In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and the show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
A set of inference rules of the form
\begin{mathpar}
	\inferrule{\Gamma_1 \\ \Gamma_2 \\ \ldots \\ \Gamma_n}{\Delta}
\end{mathpar}
where \(\Gamma_i, \Delta\) are sets of formulae and \(\lnot\left(\Gamma_1 \wedge \Gamma_2 \wedge \ldots \Gamma_n\right) \Rightarrow \lnot \Delta\) is an unconditional tautology, form a refutation calculus.
Proofs in this calculus can be expressed by closed trees of inference rules.

\subsection{Complete Inferences}
The inference rules in Figure \ref{figure:complete-inferences} form a complete inference system, by analogy with a first-order tableaux calculus without unification.
A difference and point of interest is the rule for instantiating universal quantifiers: instead of instantiating a variable with any possible term \(t\) --- an infinite space --- it is instantiated with one function symbol (or constant) at a time, quantifying over new variables as needed.
This allows for instantiating any term over multiple inference rules (effectively enumerating the Herbrand universe for the goal), but without an infinite number of possible inferences at any point.
Equality is handled by a rule rewriting classes of equal ground terms.
Both of these rules are complete yet inefficient, but both are likely to be used only a few times in order to provide enough of a ``hint'' to the oracle system for it to find a proof.

\begin{figure}
	\centering
	\begin{mathpar}
		\inferrule[contradiction]{ }{\phi, \lnot \phi, \Gamma}\and
		\inferrule[equal]{t = s, \phi\left[t/s\right], \Gamma}{t = s, \phi, \Gamma}\and
		\inferrule[implies]{\lnot \phi, \psi, \Gamma}{\phi \Rightarrow \psi, \Gamma}\and
		\inferrule[equivalent]{\lnot \phi, \lnot \psi, \Gamma\\\phi, \psi, \Gamma}{\phi \equiv \psi, \Gamma}\and
		\inferrule[conjunction]{\phi_1, \phi_2, \ldots, \phi_n, \Gamma}{\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n, \Gamma}\and
		\inferrule[disjunction]{\phi_1, \Gamma \\ \phi_2, \Gamma \\ \ldots \\ \phi_n, \Gamma}{\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n, \Gamma}\and
		\inferrule*[lab={instantiation}]{\forall x_1, x_2, \ldots x_n. \phi[f(x_1, x_2, \ldots x_n)/x], \forall x. \phi, \Gamma}{\forall x. \phi, \Gamma}\and
		\inferrule[non-empty]{\phi[k/x], \Gamma}{\forall x. \phi, \Gamma}\and
		\inferrule[exists]{\phi[k/x], \Gamma}{\exists x. \phi, \Gamma}
	\end{mathpar}
	\caption{A complete inference system for \lerna{}. Rules for negation are as usual and not shown here for brevity. In rule \textsc{instantiation}, \(f\) is a function symbol of arity \(n\) in the conclusion's signature and \(x_1 \ldots x_n\) are fresh for the conclusion. In rules \textsc{non-empty} and \textsc{exists}, \(k\) is fresh for the conclusion. \(\phi[t/s]\) is a capture-avoiding substitution replacing \(t\) for \(s\) in \(\phi\).}
	\label{figure:complete-inferences}
\end{figure}

\subsection{Weakening}
A weakening rule is an important part of \lerna{}'s calculus, since the \textsc{instantiation} and \textsc{equal} rule can produce a large number of formulae, some of which must be removed to help the oracle to prove the goal.
Each application of the rule removes some amount of information from the goal in order to simplify it --- this is sound and corresponds to removing an axiom from proof search.
The rule is merely
\begin{mathpar}
	\inferrule[weaken]{\Gamma}{\phi, \Gamma}
\end{mathpar}

\subsection{Simplifications}
Before each inferred goal is added to proof search, it is simplified with a simplification function mapping a goal to a possibly-simpler goal.
This allows for efficient removal of tedious inferences such as double-negation elimination and generally reduces the search space.
A set of example simplification rules are shown in figure \ref{figure:simplification-rules}.

\begin{figure}
	\centering
	\begin{mathpar}
		\inferrule[double-negation]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
		\inferrule[conjunction-associative]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
		\inferrule[disjunction-propagation]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
		\inferrule[reflexivity]{\top, \Gamma}{t = t, \Gamma}\and
		\inferrule[free]{\phi, \Gamma}{\forall x. \phi, \Gamma}
	\end{mathpar}
	\caption{Some simplification rules implemented in \lerna{}. In rule \textsc{free}, \(x\) is free in \(\phi\). Several other rules are implemented.}
	\label{figure:simplification-rules}
\end{figure}

\section{Oracle}
\label{section:oracle}
One problem with the calculus as described is that proofs can be quite lengthy, even if the goal is relatively trivial.
To rectify the situation, new goals generated by ongoing proof search are enqueued for attempted proof by an existing \emph{oracle} ATP system, as described in Section \ref{section:Design}.

>>>>

\documentclass{llncs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{mathpartir}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{wrapfig}

\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neural, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
	Giles Reger
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}, \email{giles.reger@manchester.ac.uk}
}
\maketitle

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Recent advances~ in neural network systems allow for processing graph-structured data in a neural context.
Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs), suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
However, in practice there are several implementation issues~\cite{lerna-discussion} which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
\begin{enumerate}
	\item Proof state in such systems may be very large, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate.
	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
\end{enumerate}
%
Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures. Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} takes an alternative step toward useful neural automatic theorem provers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}

We assume basic familiarity with first-order logic, theorem proving, and  neural networks~\cite{neural-survey}.

\subsection{Logic and Theorem Proving}

\paragraph{First-Order Logic}

\lerna{} works with formulas in standard first-order logic with equality. Terms $t$ and formulas $\phi$ are recursively defined as follows
\[
\begin{array}{ll}
t = & x \mid f\left(t_1, \ldots, t_n\right) \mid c \\
\phi  =  & \top \mid \bot \mid t_1 = t_2 \mid p\left(t_1, \ldots, t_n\right) \mid \phi_1 \Rightarrow \phi_2 \mid \phi_1 \equiv \phi_2\mid \neg \phi \mid \\ &   \phi_1 \wedge \ldots \wedge \phi_n \mid \phi_1  \vee \ldots \vee \phi_n \mid \forall x. \phi \mid \exists x. \phi
\end{array}
\]
where $x$ is a variable, $f$ is a $n$-ary function symbol, $c$ is a constant, and $p$ is a $n$-ary predicate symbol. Their meaning is defined as usual.

%A formula is either true \(\top\), false \(\bot\), an equality between terms \(t_1 = t_2\), a predicate applied to terms \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol, a negated formula \(\lnot \phi\), an impliciation \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
%Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
%Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).

\paragraph{Modern Automatic Theorem Provers}
An automated theorem prover (ATP) is a system able to automatically establish whether a formula (in first-order logic) is \emph{satisfiable} or \emph{unsatisfiable}; although, given the undecidability of this problem, ATPs may also return \emph{unknown}. Both saturation-based provers (e.g. E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}) and SMT solvers (e.g. CVC4~\cite{CVC4} and Z3~\cite{Z3}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
%Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
%Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.

\subsection{Machine Learning and Theorem Proving}
Despite the efficiency of modern ATP systems, they can still spend time exploring areas that a human mathematician would discard, and tuning such systems is, in general, extremely difficult~\cite{portfolio}.
This has led to the application of machine-learning techniques, with the eventual aim of an ``intelligent'' theorem prover able to learn from past experience to develop an intuition, discard uninteresting search space, and tune performance in a more principled way.

Previous work has focussed on premise selection~\cite{DeepMath,MaSh,MaLARea,formula-graph}, static strategy selection~\cite{Bridge,MaLeS,E-MaLeS}, dynamic (run-time) strategy selection~\cite{dynamic-strategy-priority} and more recently, direct proof guidance~\cite{MaLeCoP,FEMaLeCoP,rlCoP,ENIGMA,DNGPS}.
Proof guidance typically involves a form of machine-learned \emph{heuristic} which biases proof search in some way, allowing the prover to avoid parts of the search space deemed uninteresting by the heuristic.

Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.

\paragraph{rlCoP} The rlCoP system~\cite{rlCoP} is a connection-based reinforcement-learning prover which is not presently neurally-guided, but takes a similar approach to that taken in this paper and achieves impressive results.

\paragraph{Neural Networks for Formulae}
Neural networks are well-known tools for supervised learning~\cite{neural-survey}, and combined with trainable convolution/pooling operators are suitable for processing large-scale data such as images~\cite{cnn}.

@@

\paragraph{The MPTP Problem Set}
For training and evaluation purposes a set of problems exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
Urban et al.~\cite{rlCoP} took a subset\footnote{\url{https://github.com/JUrban/deepmath/blob/master/M2k_list}} of the large \mizarlarge{} problem set (containing 32,524 problems) and called it \mizarsmall{} (containing 2004 problems).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Design}
\label{section:Design}
In order to achieve the goal of a neural theorem prover without the disadvantages associated with neural approaches, a new design of theorem prover is required.
Popular calculi used in existing ATPs tend to be unfriendly to neural guidance.
For such a system, we require the following from the calculus:
\begin{enumerate}
        \item \emph{Proof state must be small.} Attempting to evaluate large proof states structurally requires a lot of computation and resources. Saturation-based provers can have very large proof states, for example.
        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms operate more efficiently in batches. Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space.%, without e.g. unifiers. 
        Otherwise, the learning system is trying to learn while blind to the context of the search.
	\item \emph{Subgoals must be intelligible.} Adding ``noise'' such as clausification obscures the original intuition behind a goal, at least for human observers. While this is not necessarily the case for machine-learning algorithms, it seems likely that removing structure and adding artefacts will reduce model performance.
\end{enumerate}
We therefore implement a refutation prover based on a first-order tableaux calculus without unification, on non-clausal formulae.
Each goal in this case is the set of formulae present on the tableau branch.
In this context, proof state is small (only the current branch), evaluation of states is possible in parallel, each branch is independent and contains all information required, and all available structure from the original problem is kept.

\subsection{Search}
In the calculus (see Section \ref{section:Calculus}) for this prover, there are two branching factors: each goal has a set of possible inferences, and each inference contains a set of possible sub-goals.
To prove a goal, at least one inference must be proved.
To prove an inference, all the inferences' sub-goals must be proved.
A simple optimisation observation is that sub-goals might well be shared between inferences from all over the search space, so search becomes a directed acyclic graph, alternating between goals and inferences (illustrated in Fig.~\ref{fig:search}).

%\begin{figure}[t]
\begin{wrapfigure}[12]{r}{0.5\textwidth} 
\vspace{-26pt}
	\centering
	\includegraphics[width=0.9\linewidth]{search2.pdf}
	\caption{Search in the \lerna{} system, showing shared sub-goals.\label{fig:search}}
\end{wrapfigure}	
%\end{figure}

Now the search graph can be explored: in each step, a leaf (goal) node is selected for expansion, and all resulting inferences and sub-goals are added to the graph.
If a goal has no possible inferences, it is satisfiable and can be removed from the search space.
On the other hand, if a goal is trivial (i.e. contains a contradiction), it is unsatisfiable and can be marked as proven.
This idea is lifted to inferences: if an inference contains any satisfiable sub-goal, it too is satisfiable, whereas if an inference contains all unsatisfiable goals, it is unsatisfiable.
Proof search continues until the timeout is reached or the root goal is shown to be (un-)satisfiable.
In order to dispatch trivial sub-goals quickly, an existing fast \emph{oracle} ATP is used (see Section \ref{section:oracle}).
This may mark goals as (un-)satisfiable, at which point no further exploration is required.

@@

\subsection{Architecture and Prototype Implementation}
The system aims to consume all available CPU and GPU resources as efficiently as possible.
To that end, proof search is asynchronous: the search algorithm generates new sub-goals, which are placed on two separate queues: one for the oracle ATP, another for heuristic evaluation.
Proof search then continues elsewhere, while the oracle ATP is called in parallel on each sub-goal (consuming all available CPU) while the heuristic consumes batches of subgoals, efficiently utilising the available computational resource.
As information flows backwards from these processes, the search process updates its information about a given sub-goal and propagates that information upwards to the sub-goal's parent inferences, to influence future proof search: see Fig. \ref{figure:architecture}.

\begin{figure}[t]
	\centering
	\includegraphics[width=0.8\linewidth]{architecture}
	\caption{Illustrating information-flow in the system.}
	\label{figure:architecture}
\end{figure}

%\subsection{}

The prototype implementation (minus the heuristic) %was built in the Rust programming language.% was chosen to implement the main prover , for its memory and thread-safety guarantees, as well as its support for managing memory allocation and algebraic datatypes.
%The implementation 
is currently just under 3,000 lines of Rust, not including \z3{}, the TPTP format parser or the implementation of perfect sharing.
Python 3 was used for the heuristic due to the large number of libraries available for neural network implementation in Python.
The heuristic is implemented as a server, communicating with the main prover via a TCP socket.
In principle this allows for the heuristic to be a shared resource with a centralised heuristic server, or a load-balanced cluster.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Calculus}
\label{section:Calculus}
The proof calculus used in the above architecture may be extremely general: in fact, any function from goals to a finite set of possible inferences (themselves finite sets of sub-goals) will suffice, as long as each goal remains independent of any other such that the heuristic function can process all available information.  If the inference system is complete, there are no additional constraints such as orderings or fairness to ensure the completeness of the prover, as the balanced search algorithm (see Section \ref{section:Design}) will ensure this.

\lerna{} presently implements a refutation tableaux calculus~\cite{handbook-of-automated-reasoning} without unification.
The calculus described is deliberately na\"ive in order to easily satisfy the design constraints given above, but may be replaced by a stronger calculus in the future. 
A na\"ive calculus is not necessarily a problem as the heuristic should select promising areas to explore and ignore uninteresting sub-goals.
However, a more efficient calculus would improve performance where the heuristic fails.

\paragraph{Refutation Tableaux}
In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and the show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
A set of inference rules of the form
\begin{mathpar}
	\inferrule{\Gamma_1 \\ \Gamma_2 \\ \ldots \\ \Gamma_n}{\Delta}
\end{mathpar}
where \(\Gamma_i, \Delta\) are sets of formulae and \(\lnot\left(\Gamma_1 \wedge \Gamma_2 \wedge \ldots \Gamma_n\right) \Rightarrow \lnot \Delta\) is an unconditional tautology, form a refutation calculus.
Proofs in this calculus can be expressed by closed trees of inference rules.

\begin{figure}[t]
	\centering
	\begin{mathpar}
		\inferrule[contradiction]{ }{\phi, \lnot \phi, \Gamma}\and
		\inferrule[equal]{t = s, \phi\left[t/s\right], \Gamma}{t = s, \phi, \Gamma}\and
		\inferrule[implies]{\lnot \phi, \psi, \Gamma}{\phi \Rightarrow \psi, \Gamma}\and
		\inferrule[equivalent]{\lnot \phi, \lnot \psi, \Gamma\\\phi, \psi, \Gamma}{\phi \equiv \psi, \Gamma}\and
		\inferrule[conjunction]{\phi_1, \phi_2, \ldots, \phi_n, \Gamma}{\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n, \Gamma}\and
		\inferrule[disjunction]{\phi_1, \Gamma \\ \phi_2, \Gamma \\ \ldots \\ \phi_n, \Gamma}{\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n, \Gamma}\and
		\inferrule*[lab={instantiation}]{\forall x_1, x_2, \ldots x_n. \phi[f(x_1, x_2, \ldots x_n)/x], \forall x. \phi, \Gamma}{\forall x. \phi, \Gamma}\and
		\inferrule[non-empty]{\phi[k/x], \Gamma}{\forall x. \phi, \Gamma}\and
		\inferrule[exists]{\phi[k/x], \Gamma}{\exists x. \phi, \Gamma}
	\end{mathpar}
	\caption{A complete inference system for \lerna{}. Rules for negation are as usual and not shown here for brevity. In rule \textsc{instantiation}, \(f\) is a function symbol of arity \(n\) in the conclusion's signature and \(x_1 \ldots x_n\) are fresh for the conclusion. In rules \textsc{non-empty} and \textsc{exists}, \(k\) is fresh for the conclusion. \(\phi[t/s]\) is a capture-avoiding substitution replacing \(t\) for \(s\) in \(\phi\).}
	\label{figure:complete-inferences}
\end{figure}

\paragraph{Complete Inferences}
The inference rules in Fig. \ref{figure:complete-inferences} form a complete inference system, by analogy with a first-order tableaux calculus without unification.
A difference and point of interest is the rule for instantiating universal quantifiers: instead of instantiating a variable with any possible term \(t\) --- an infinite space --- it is instantiated with one function symbol (or constant) at a time, quantifying over new variables as needed.
This allows for instantiating any term over multiple inference rules (effectively enumerating the Herbrand universe for the goal), but without an infinite number of possible inferences at any point.
Equality is handled by a rule rewriting classes of equal ground terms.
Both of these rules are complete yet inefficient, but both are likely to be used only a few times in order to provide enough of a ``hint'' to the oracle system for it to find a proof.

\paragraph{Weakening}
A weakening rule is an important part of \lerna{}'s calculus, since the \textsc{instantiation} and \textsc{equal} rule can produce a large number of formulae, some of which must be removed to help the oracle to prove the goal.
Each application of the rule removes some amount of information from the goal in order to simplify it --- this is sound and corresponds to removing an axiom from proof search.
The rule is merely\vspace{-0.5em}
\begin{mathpar}
	\inferrule[weaken]{\Gamma}{\phi, \Gamma}
\end{mathpar}

\paragraph{Simplifications}
Before each inferred goal is added to proof search, it is simplified, % with a simplification function mapping a goal to a possibly-simpler goal.
%This allows for efficient removal of 
removing 
tedious inferences such as double-negation elimination and generally %reduces 
reducing 
the search space.
Fig. \ref{figure:simplification-rules} gives example simplification rules.

\begin{figure}[t]
	\centering
	\begin{mathpar}
%		\inferrule[double-negation]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
%		\inferrule[conjunction-associative]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
%		\inferrule[disjunction-propagation]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
%		\inferrule[reflexivity]{\top, \Gamma}{t = t, \Gamma}\and
%		\inferrule[free]{\phi, \Gamma}{\forall x. \phi, \Gamma}
		\inferrule[double-neg]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
		\inferrule[conj-assoc]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
		\inferrule[disj-prop]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
		\inferrule[refl]{\top, \Gamma}{t = t, \Gamma}\and
		\inferrule[free]{\phi, \Gamma}{\forall x. \phi, \Gamma}
	\end{mathpar}
	\caption{Some simplification rules implemented in \lerna{}. In rule \textsc{free}, \(x\) is free in \(\phi\). Several other rules are implemented.}
	\label{figure:simplification-rules}
\end{figure}
